Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)

### 2023-05-23 doc/manuscript/ESSDmanuscript.Rmd
incorporate Sandro's comments (send in Word doc)
plus  a few other changes, including title

#39

---
title: Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'essd')`"
author:
  - given_name: Kristina J.
    surname: Anderson-Teixeira
    affiliation: "1,2 *"
    email: teixeirak@si.edu
    corresponding: true
  - given_name: Valentine
    surname: Herrmann
    affiliation: "1"
  - given_name: Madison 
    surname: Williams
    affiliation: "1"
  - given_name: Teagan
    surname: Rogers
    affiliation: "1"
  - given_name: Rebecca
    surname: Banbury Morgan
    affiliation: "1,3" 
  - given_name: Ben 
    surname: Bond-Lamberty
    affiliation: "4" 
  - given_name: Susan 
    surname: Cook-Patton
    affiliation: "5" 
  - given_name: Camille
    surname: Piponiot
    affiliation: "6"

@@

**Abstract.**
Forests are critical for climate change mitigation and constitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhoouse gas emissions and removals.
We begin by reconciling terminology and mapping ForC fields into EFDB.
This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
This distribution of records is generally reflected in the subset of records that have been submitted to EFDB to date.
In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C accounting.

@@

Forests are critical to management of atmospheric concentrations of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 Gt C yr^-1^ [all biomes with trees, @xu_changes_2021] to ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al. Table 2--> [forests only, @harris_global_2021].
This has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al. Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
Going into the future, the fate of this important CO~2~ sink is highly uncertain, depending both upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020;@hammond_global_2022], and on human conservation, restoration, and management of forests [@ipcc_climate_2019;@ipcc_climate_2022].

@@

The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2006_2006;@ipcc_2019_2019].
Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
The IPCC inventory guidelines include specific instructions for accounting for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
A tiered approach to accounting is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances.
For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default values underestimated C sequestration by 32% on average and failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020]. 
In addition, it was revealed that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
This rapid evolution of scientific information on C cycling in forests is valuable for informing climate change mitigation efforts but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning. 
Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.

To improve data accessibility for C accounting, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
In the case of forests, emission factors include C stocks, net increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].

The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
As such, ForC is positioned to improve forest C accounting through the transfer of data to EFDB.
The purpose of this publication is to document that process and provide recommendations for future improvements.

Here, we
(1) review IPCC methods and definitions for forest C accounting in the context of typical forest C estimation methodologies;
(2) describe mapping of ForC to IPCC's EFDB;
(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data transfer to EFDB;
(4) summarize the data in ForC that's relevant to EFDB and records that have been transferred to date; and
(5) provide recommendations for improving data collection, analysis, database, and accounting.

@@

Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
As noted above, live fine roots may be included in litter when difficult to separate empirically.
The definition includes the entire O horizion, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
This definition contrasts with empirical studies that focus on aboveground litter, often including only the OL layer in the definition of litter, and do not always specify the components included.
Similar to dead wood, litter is poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->

@@

Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C accounting under IPCC guidelines. 
However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.

@@

# Transfer of data from ForC to EFDB

To transfer complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
Criteria for data transfer were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been sent to EFDB.
Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for inclusion in EFDB. 
Complete records needed to be reviewed by the EFDB editorial board prior to posting in the database -- a process that lags behind records transfer and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.

@@

With input from the IPCC's Technical Support Unit, we reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into EFDB (Fig. `r fig_variable_mapping`).
For each C pool (Table `r table_pools`), we identified variables representing organic matter or C stocks, stock changes (a.k.a. "net annual increments" by IPCC, "increments" in ForC), gains (a.k.a. "gross annual increments" by IPCC, "fluxes" in ForC), and losses ("fluxes" in ForC).
As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC. 
It it important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
For example, ForC records of biomass and dead wood vary in the minimum stem diameter censused, such that some records would match the IPCC criteria whereas others would not.
Information on minimum diameters censused and other important sources of methodological variation are recorded as covariates in ForC and mapped into the EFDB field *Other Properties* (`r table_ForCfieldmapping`).
Details on the mapping of ForC variables to EFDB -- including associated covariates, IPCC pools (Table `r table_pools`) and relevant equations [@ipcc_2006_2006] -- are documented in the file ForC_variables_mapping.csv in the GitHub repository associated with this publication IPCC-EFDB-integration repository in ForC-db organization ([https://github.com/forc-db/IPCC-EFDB-integration](https://github.com/forc-db/IPCC-EFDB-integration)).

@@

Classification into sub-categories was dependent upon stand age and site history (section `r  section_land_use_categories`).
For Forest Land ≥ 20 years old or of unknown (relatively mature) age, or Forest Land < 20 years old that was forest prior to a stand-clearing disturbance, the past land-use category was Forest Land, making the sub-category "Forest Land Remaining Forest land".
For forests <20 years old with history including cultivation/ tillage or grazing, past land-use categories were Cropland and Grassland, respectively, making land-use subcategories were "Cropland converted to Forest Land" and "Cropland converted to Forest Land", respectively.
For forests <20 years old with unspecified previous agricultural use, we assigned the sub-category "Land Converted to Forest land".
Forests <20 years old with unknown land use prior to the study date were simply classified as "Forest Land".
The same logic was applied for savannas, but including both Forest Land and Grassland as potentially relevant categories.

Given the lack of public information needed to determine whether lands are classified as mangaged [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not transfer any classification of land management status from ForC to the EFDB.
However, we do provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.

@@

As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the the `r n_relevant_variables` variables relevant to EFDB (Fig. `r fig_variable_mapping`).
These records were distributed across all forested continents and ecozones, albeit unevenly (Fig. `r fig_map`).
The largest number of records came from Asia, followed by North America, South America, and Europe, with relatively few records from Africa, Australia, and Oceania (Fig. `r fig_histograms`c).
Categorized by FAO ecozone, the greatest numbers of records came from subtropical humid forests, temperate mountain systems, and tropical rain forests, each with >2,000 independent records (Fig. `r fig_histograms`b). 
Boreal coniferous forests, temperate continental forests, subtropical mountain systems, and tropical moist deciduous forests had >1,000 independent records each, while other ecozones all had <1,000 records.
The most widely represented forest type was needleleaf evergreen, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`a).
In terms of stand age, the most represented age class was 20-100 years, followed by <20 years and then >100 years (Fig. `r fig_histograms`d).

@@

## Data transfers to EFDB

As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed or added `r n_records_reviewed_or_added` EFDB-relevant records, `r n_records_sent_to_EFDB` records of which were sent to EFDB, and `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs. `r fig_map`-`r fig_histograms`, Table `r table_variables`).
The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records reviewed and those sent to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
The discrepancy between the number of records sent and that posted to EFDB is primarily attributable to the time required for the IPCC to review the records, and it is not expected that many -- if any -- records will be rejected.

@@

There is vast potential to expand forest C data in EFDB by completing the process of reviewing and sending data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been sent to EFDB.
Although this process requires manual review of records, the addition of new records to EFDB is hugely facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily transferred to EFDB using the system developed here.
Future efforts to review studies for transfer should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs. `r fig_map`-`r fig_histograms`, Table `r table_variables`), to records from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), to the variables most needed by EFDB users, and to the more contemporary records.

@@

Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for forest C accounting.
Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023].
Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.
Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@tifafi_large_2018].
Significant investment in ground-based forest research will be critical to filling these gaps.

@@

The ForC database contains large numbers of records that could potentially be useful for C accounting under IPCC guidelines.
Here we have developed a framework for transferring these records to the EFDB, making them more accessible for forest C accounting under IPCC guidelines [@ipcc_2006_2006;@ipcc_2019_2019].
As of `r format(Sys.time(), '%B %d, %Y')`, we have submitted `r n_records_sent_to_EFDB` records to EFDB.
Although this represents just `r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables*100,0)`% of relevant records in ForC, it substantially increases the number of forest land records in EFDB.
The records transferred to EFDB and present in ForC are very unevenly distributed across variables, regions, and forest types (Figs. `r fig_map`-`r fig_histograms`, Table `r table_variables`), reflecting broader patterns in allocation of research effort.

>>>>

---
title: Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'essd')`"
author:
  - given_name: Kristina J.
    surname: Anderson-Teixeira
    affiliation: "1,2 *"
    email: teixeirak@si.edu
    corresponding: true
  - given_name: Valentine
    surname: Herrmann
    affiliation: "1"
  - given_name: Madison 
    surname: Williams
    affiliation: "1"
  - given_name: Teagan
    surname: Rogers
    affiliation: "1"
  - given_name: Rebecca
    surname: Banbury Morgan
    affiliation: "1,3" 
  - given_name: Ben 
    surname: Bond-Lamberty
    affiliation: "4" 
  - given_name: Susan 
    surname: Cook-Patton
    affiliation: "5" 
  - given_name: Camille
    surname: Piponiot
    affiliation: "6"

@@

**Abstract.**
Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhouse gas emissions and removals.
We begin by reconciling terminology and mapping ForC fields into EFDB.
This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
This distribution of records is generally reflected in the subset of records that have been submitted to EFDB to date.
In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C accounting.

@@

Forests are critical to management of the atmospheric concentration of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 Gt C yr^-1^ [all biomes with trees, @xu_changes_2021] to ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al. Table 2--> [forests only, @harris_global_2021].
This has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al. Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and associated climate change.
Going into the future, the fate of this important CO~2~ sink is highly uncertain, depending both upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020;@hammond_global_2022], and on human conservation, restoration, and management of forests [@ipcc_climate_2019;@ipcc_climate_2022].

@@

The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2006_2006;@ipcc_2019_2019].
Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
The IPCC inventory guidelines include specific instructions for accounting for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
A tiered approach to accounting is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances and datasets grow and become more widely accessible.
For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default values underestimated C sequestration by 32% on average and failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020]. 
In addition, it was revealed that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
This rapid evolution of scientific information on C cycling in forests is valuable for informing climate change mitigation efforts but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning. 
Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.

To improve data accessibility for preparing greenhouse gas estimates, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
In the case of forests, emission factors include C stocks, net increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].

The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.
The purpose of this publication is to document that process and provide recommendations for future improvements.

Here, we
(1) review IPCC methods and definitions applied to estimate CO2 emissions and removals from forest in the context of typical forest C estimation methodologies;
(2) describe mapping of ForC to IPCC's EFDB;
(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data transfer to EFDB;
(4) summarize the data in ForC that's relevant to EFDB and records that have been submitted to date; and
(5) provide recommendations for improving data collection, analysis, database, and accounting.

@@

Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
As noted above, live fine roots may be included in litter when difficult to separate empirically.
The definition includes the entire O horizion, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
This definition contrasts with empirical studies that focus on aboveground litter, often including only the OL layer in the definition of litter, and do not always specify the components included.
Similar to dead wood, litter is poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->

@@

Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C stock change calculations under the Paris Agreement. 
However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C inventories under IPCC guidelines.
To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.

@@

# Submission of ForC data to EFDB

To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
Criteria for data submission were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been sent to EFDB.
Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for inclusion in EFDB. 
Complete records needed to be reviewed by the EFDB editorial board prior to posting in the database -- a process that lags behind records transfer and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.

@@

With input from the IPCC's Technical Support Unit, we reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into EFDB (Fig. `r fig_variable_mapping`).
For each C pool (Table `r table_pools`), we identified variables representing organic matter or C stocks, stock changes (a.k.a. "net annual increments" by IPCC, "increments" in ForC), gains (a.k.a. "gross annual increments" by IPCC, "fluxes" in ForC), and losses ("fluxes" in ForC).
As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC. 
It is important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
For example, ForC records of biomass and dead wood vary in the minimum stem diameter censused, such that some records would match the IPCC criteria whereas others would not.
Information on minimum diameters censused and other important sources of methodological variation are recorded as covariates in ForC and mapped into the EFDB field *Other Properties* (`r table_ForCfieldmapping`).
Details on the mapping of ForC variables to EFDB -- including associated covariates, IPCC pools (Table `r table_pools`) and relevant equations [@ipcc_2006_2006] -- are documented in the file ForC_variables_mapping.csv in the GitHub repository associated with this publication IPCC-EFDB-integration repository in ForC-db organization ([https://github.com/forc-db/IPCC-EFDB-integration](https://github.com/forc-db/IPCC-EFDB-integration)).

@@

Classification into sub-categories was dependent upon stand age and site history (section `r  section_land_use_categories`).
For Forest Land ≥ 20 years old or of unknown (relatively mature) age, or Forest Land < 20 years old that was forest prior to a stand-clearing disturbance, the past land-use category was Forest Land, making the sub-category "Forest Land Remaining Forest land".
For forests <20 years old with history including cultivation/ tillage or grazing, past land-use categories were Cropland and Grassland, respectively, making land-use subcategories were "Cropland converted to Forest Land" and "Grassland converted to Forest Land", respectively.
For forests <20 years old with unspecified previous agricultural use, we assigned the sub-category "Land Converted to Forest land".
Forests <20 years old with unknown land use prior to the study date were simply classified as "Forest Land".
The same logic was applied for savannas, but including both Forest Land and Grassland as potentially relevant categories.

Given the lack of public information needed to determine whether lands are classified as managed [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not transfer any classification of land management status from ForC to the EFDB.
However, we do provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.

@@

As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the `r n_relevant_variables` variables relevant to EFDB (Fig. `r fig_variable_mapping`).
These records were distributed across all forested continents and ecozones, albeit unevenly (Fig. `r fig_map`).
The largest number of records came from Asia, followed by North America, South America, and Europe, with relatively few records from Africa, Australia, and Oceania (Fig. `r fig_histograms`c).
Categorized by FAO ecozone, the greatest numbers of records came from subtropical humid forests, temperate mountain systems, and tropical rain forests, each with >2,000 independent records (Fig. `r fig_histograms`b). 
Boreal coniferous forests, temperate continental forests, subtropical mountain systems, and tropical moist deciduous forests had >1,000 independent records each, while other ecozones all had <1,000 records.
The most widely represented forest type was needleleaf evergreen, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`a).
In terms of stand age, the most represented age class was 20-100 years, followed by <20 years and then >100 years (Fig. `r fig_histograms`d).

@@

## Data submissions to EFDB

As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed or added `r n_records_reviewed_or_added` EFDB-relevant records, `r n_records_sent_to_EFDB` records of which were submitted to EFDB, and `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs. `r fig_map`-`r fig_histograms`, Table `r table_variables`).
The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records reviewed and those sent to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
The discrepancy between the number of records sent and that posted to EFDB is primarily attributable to the time required for the IPCC to review the records, and it is not expected that many -- if any -- records will be rejected.

@@

There is vast potential to expand forest C data in EFDB by completing the process of reviewing and submitting data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been submitted to EFDB.
Although this process requires manual review of records, the submission of new records to EFDB is hugely facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily prepared for submission to EFDB using the system developed here.
Future efforts to review studies for transfer should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs. `r fig_map`-`r fig_histograms`, Table `r table_variables`), to records from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), to the variables most needed by EFDB users, and to the more contemporary records.

@@

Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for estimating forest C stock chnages.
Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023].
Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.
Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@tifafi_large_2018].
Significant investment in ground-based forest research will be critical to filling these gaps.

@@

The ForC database contains large numbers of records that could potentially be useful for estimating C stock changes applying methodological guidance provided by the IPCC.
Here we have developed a framework for submitting these records to the EFDB, thus making those data more accessible for reporting CO~2~ emissions and removals from forest land consistent with good practice in the IPCC guidelines  [@ipcc_2006_2006;@ipcc_2019_2019].
As of `r format(Sys.time(), '%B %d, %Y')`, we have submitted `r n_records_sent_to_EFDB` records to EFDB.
Although this represents just `r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables*100,0)`% of relevant records in ForC, it substantially increases the number of forest land records in EFDB.
The records submitted to EFDB and present in ForC are very unevenly distributed across variables, regions, and forest types (Figs. `r fig_map`-`r fig_histograms`, Table `r table_variables`), reflecting broader patterns in allocation of research effort.
